Bayesian Rate Estimation

### 2012-11-01 many.tex
curt added subsection on bad behavior of Bayesian rate estimate when there is one very loud outlier

Consider the following model problem.
% in which the distribution of noise events is Gaussian.
Let $p(\rho)\Delta\rho$ be the probability that data yields a putative event with SNR in the
range $[\rho - \Delta\rho/2, \rho + \Delta\rho/2]$  We write $p = p_n + p_s$, where $p_n$ is
the distribution of noise events and $p_s$ is the distribution of events from actual astrophysical sources.   Both $p_n$ and $p_s$ are the product of a rate (of noise or true events) and the observation time.  Here we will assume that the noise distribution is Gaussian, so that $p_n$
has the general form
\be 
p_n(\rho) = \Lambda_n {\rm exp}\big[-\rho^2/2\big] \, .
\ee
\noindent
We find it useful to define $\rho_1$ as the SNR such that a data set will have on average
a single noise event louder than $\rho_1$; i.e., such that
\be
\int_{\rho_1}^{\infty} P_n(\rho)\, d\rho  = 1 \, .
\ee
For this example we will assume that the spatial distribution of true events can be approximated as uniform in Euclidean space. Then we can write the full $p(\rho)$ as
%and that the spatial distribution of true events can be approximated as uniform in Euclidean %space. Then $p(\rho)$ has the general form  
\be\label{eq:ptot}
p(\rho) = \big[\sqrt{\pi/2} {\rm erfc}(\rho_1/\sqrt{2})\big]^{-1} \ exp[-\rho^2/2] + 3 R_t \frac{\rho_1^3}{\rho^4} \, .
\ee
\noindent where $R_t$ is the mean number of true events with $\rho > \rho_1$.
%What matters for the following analysis is the rate times the observation time.

For any pair $(\rho_1, R_t)$, it is straightforward to construct random event lists drawn from the corresponding $p(\rho)$, and straightforward to apply a threshold by "throwing away" all events with $\rho$ less than some threshold value $\rho_{th}$.  For any thresholded event list, we use
Eqs. ..... to construct the probability density $p(R)$ (except that in this example, instead of the
Jeffrey's prior, we assume a "hat-shaped" prior; see below).
% that is uniform in some interval and zero elsewhere; see below).  
For that event list, we define $\Delta R$ by
\be
(\Delta R_B)^2 = \int{(R - R_B)^2 \, p(R)\, dR }\, .
\ee
%To find the average behavior, we do many draws of $R_t$ and corresponding data sets.

Figs.~\ref{fig:delRb} --\ref{fig:Rt10}  illustrate how $\Delta R$ varies with the threshold value $\rho_{th}$.  In all cases we assumed that $\rho_1 = 8$.
%
For Fig.~\ref{fig:delR1}, we adopted a flat prior on $R$ between $1$ and $10$ (i.e, we took
$p(R) = 1/9$  for $1 < R <10$, and  zero outside that interval).  
% To create Fig.~\ref{fig:delR1}, 
We drew $10^3$ 
%\curt{Curt needs to check this} 
random realizations $(R_t, data)$, where $R_t$ was drawn from that flat prior and the data--a list of $\rho$ values-- was drawn from the distribution $p(data|R_t)$ given by Eq.~\ref{eq:ptot} .  For any given threshold value, $\rho_{th}$, we created an event list by taking the subset of the $\rho$ values
% in the data set that are 
greater than $\rho_{th}$  For each data set, we used a sequence of increasing threshold  values $\rho_{th}$ to create a corresponding sequence of (ever shorter)  event lists.  For each event list, we calculated $\Delta R$ and $R_B - R_t$.
%and take the mean over all realizations.  
As a code check, we verified that the mean of $R_B - R_t$ (over all realizations) was consistent with zero for all $\rho_{th}$.  Fig.~\ref{fig:delR1} displays both the mean value of $\Delta R$ and the mean number of noise events (as opposed to actual signals) as a function of $\rho_{th}$. 
Note the relative insensitivity of  $\Delta R_t$ to $\rho_{th}$:  as the mean
number of noise events on the event list shrinks from $8$ to $\sim 10^{-6}$, 
the average $\Delta R_B$ increases only $\sim 14\%$, from 0.42to 0.48.

@@

Figs.~\ref{fig:Rt1} --\ref{fig:Rt10} were constructed similarly, but with the following differences.  First, we took the prior 
$p(R)$ to be $p(R) \propto 1/R$ for  $0.03 < R < 30$, and zero outside that interval (i.e., $p(R)$ is the Jeffries prior, but restricted to some range).   Second, we regarded $R_t$ as fixed (as of course it would be in Nature), and only varied the data--randomly drawing $10^3$ realizations from $p(data|R_t)$.  For each realization we calculated $\Delta R_B$ and $(R_B - R_t)$, and then we calculated the mean over all realizations.  The mean of $(R_B - R_t)$ was generally {\it not} zero, since the
prior biases the results.  (Of course, if one {\it updated} the prior after each realization, the 
bias would decrease towards zero, but that is not what we did: we evaluated each realization with the same Jeffries prior.)

The number of noise events is the same as for the first case, shown in Fig.~\ref{fig:delR}.  Note that again, for all three values of $R_t$, the mean uncertainty $\Delta R_B$ increases by only $\sim 20\%$  as the mean number of noise events on the event list shrinks from $8$ to $\sim 10^{-6}$.  And as one would expect, the relative uncertainty shrinks as $R_t$ increases.

@@

>>>>

Consider the following model problem.
% in which the distribution of noise events is Gaussian.
Let $p(\rho)\Delta\rho$ be the probability that data yields a putative event with SNR in the
range $[\rho - \Delta\rho/2, \rho + \Delta\rho/2]$  We write $p = p_n + p_s$, where $p_n$ is
the distribution of noise events and $p_s$ is the distribution of events from actual astrophysical sources.   Both $p_n$ and $p_s$ are the product of a rate (of noise or true events) and the observation time.  Here we will assume that the noise distribution is Gaussian, so that $p_n$
has the general form
\be 
p_n(\rho) = \Lambda_n {\rm exp}\big[-\rho^2/2\big] \, .
\ee
\noindent
We find it useful to define $\rho_1$ as the SNR such that a data set will have on average
a single noise event louder than $\rho_1$; i.e., such that
\be
\int_{\rho_1}^{\infty} P_n(\rho)\, d\rho  = 1 \, .
\ee
For this example we will assume that the spatial distribution of true events can be approximated as uniform in Euclidean space, so 
\be\label{eq:ps}
p_s(\rho) = 3 R_t \frac{\rho_1^3}{\rho^4} \, ,
\ee
\noindent where $R_t$ is the mean number of true events with $\rho > \rho_1$.
Then we can write the full $p(\rho)$ as
%and that the spatial distribution of true events can be approximated as uniform in %Euclidean %space. Then $p(\rho)$ has the general form  
\be\label{eq:ptot}
p(\rho) = \big[\sqrt{\pi/2} {\rm erfc}(\rho_1/\sqrt{2})\big]^{-1} \, {\rm exp}[-\rho^2/2]\  + \ 3 R_t \frac{\rho_1^3}{\rho^4} \, .
\ee

%What matters for the following analysis is the rate times the observation time.

For any pair $(\rho_1, R_t)$, it is straightforward to construct random event lists drawn from the corresponding $p(\rho)$, and straightforward to apply a threshold by "throwing away" all events with $\rho$ less than some threshold value $\rho_{th}$.  For any thresholded event list, we use
Eqs.~(\curt{need to supply, from earlier in paper} ) to construct the probability density $p(R|data)$ (except that in this example, instead of the
Jeffrey's prior, we assume a "hat-shaped" prior; see below).
% that is uniform in some interval and zero elsewhere; see below).  
For that event list, we define $\Delta R_B$ by
\be
(\Delta R_B)^2 = \int{(R - R_B)^2 \, p(R|data)\, dR }\, .
\ee
%To find the average behavior, we do many draws of $R_t$ and corresponding data sets.
%
Figs.~\ref{fig:delRb} --\ref{fig:Rt10}  illustrate how $\Delta R$ varies with the threshold value $\rho_{th}$.  In all cases we assumed that $\rho_1 = 8$.
%

For Fig.~\ref{fig:delRb}, we adopted a flat prior on $R$ between $1$ and $10$ (i.e, we took
$p(R) = 1/9$  for $1 < R <10$, and  zero outside that interval).  
% To create Fig.~\ref{fig:delR1}, 
We drew $10^4$ 
random realizations $(R_t, data)$, where $R_t$ was drawn from that flat prior and the data--a list of $\rho$ values-- was drawn from the distribution $p(data|R_t)$ given by Eq.~\ref{eq:ptot} .  For any particular threshold value, $\rho_{th}$, we created an  event list by taking the subset of the $\rho$ values
% in the data set that are 
that were greater than $\rho_{th}$.  For each data set, we considered a sequence of increasing threshold  values $\rho_{th}$, using them to create a corresponding sequence of (ever shorter)  event lists.  For each event list, we calculated $\Delta R$ and $R_B - R_t$.
%and take the mean over all realizations.  
As a code check, we verified that the mean of $R_B - R_t$ (over all realizations) was consistent with zero for all $\rho_{th}$.  Fig.~\ref{fig:delR1} displays both the mean value of $\Delta R$ and the mean number of noise events (as opposed to actual signals) as a function of $\rho_{th}$. 
Note the relative insensitivity of  $\Delta R_t$ to $\rho_{th}$:  as the mean
number of noise events on the event list shrinks from $8$ to $\sim 10^{-6}$, 
the average $\Delta R_B$ increases only $\sim 14\%$, from $0.42$ to $0.48$.

@@

Figs.~\ref{fig:Rt1} --\ref{fig:Rt10} were constructed similarly, but with the following differences.  First, we took the prior 
$p(R)$ to be $p(R) \propto 1/R$ for  $0.03 < R < 30$, and zero outside that interval (i.e., $p(R)$ is the Jeffries prior, but restricted to some range).   Second, we regarded $R_t$ as fixed (as of course it would be in Nature), and only varied the data--randomly drawing $10^4$ realizations from $p(data|R_t)$.  For each realization we calculated $R_t^{-1}\,\Delta R_B$ and $(R_B - R_t)$, and then we calculated the means of these quantities over all realizations.  The mean of $(R_B - R_t)$ was generally {\it not} zero, since the
prior biases the results.  (Of course, if one {\it updated} the prior after each realization, the 
bias would decrease towards zero, but that is not what we did: we evaluated each realization with the same Jeffries prior.)

For Figs.~\ref{fig:Rt1} --\ref{fig:Rt10}, the mean number of noise events as a function of $\rho_{th}$ was the same as for the case shown in Fig.~\ref{fig:delRb}.  Note that for all three values of $R_t$, the mean uncertainty $\Delta R_B$ increases by only $\sim 20\%$  as the mean number of noise events on the event list shrinks from $8$ to $\sim 10^{-6}$. 
Thus the inference that the average $\Delta R_B$ is rather insensitive to  $\rho_{th}$ appears to be robust.
Finally, we note that for all three cases the bias is much smaller than $\Delta R_B$.

@@

\subsection{Extreme Sensitivity of the LE Rate Estimate to a Single, Unusually Loud Event}
Here we discuss a very unattractive feature of the Bayesian loudest event estimate of $R$:
a small percentage of the time it will yield a very large over-estimate.

To explain this, we will use the same model as described in the previous subsection, and we will begin with a very specific example.   Let the true rate (times the observation time) be $R_t = 1$, meaning (again) that the expected number of actual events with $\rho > \rho_1$ is one.   Then there is a $1/64$ chance ($\approx 1.6\%$) that the loudest event will have $\rho_{LE} > 4 \rho_1$.  Consider this case, and let us also assume that there are no events (noise or actual) with $\rho_1 < \rho < \rho_{LE}$.  The loudest event estimate basically "throws away" the information that there
are no events in this interval, which (we show below) leads an estimate $R_{LE} \approx 64$,
  which is $64$ times the true rate!  Now, if the true rate really were $R_t = 64$, then the expected number of 
events with $\rho > \rho_1$ would be $64$.  So in this case, the loudest event estimate 
ignores the fact that there are $\sim 56 -72$ "missing" events.  However a Bayesian estimate
with $\rho_{th}$ set to $\rho_1$ incorporates this information quite naturally, and so (correctly) yields an estimated $R$ of order one.

We now essentially repeat the above discussion, but in greater generality and with the derivations provided.   First recall that $R_B$ is given by
\be\label{eq:rb1}
R_B = \frac{  \int{ R\, p(\{\rho_i\}|R)  \, p(R)\, dR}}{  \int{ p(\{\rho_i\}|R)\, p(R)\, dR }}\, ,
\ee
\noindent
where $p(R)$ is the prior on $R$ and (and up to an overall normalization)
\be\label{eq:rb2}
p(\{\rho_i\}|R) = {\rm exp}\bigg[\sum_{i \in  list} {\rm ln}\, p(\rho_i|R) - \int_{\rho_{th}}^{\infty} p(\rho)\, d\,\rho  \bigg] \,  ,
\ee
\noindent
with $p(\rho)$ given by Eq.~().

The loudest event estimate is obtained by setting the lower integration limit in the last term on the rhs of Eq.~(\ref{eq:rb2}) to $\rho_{th} = \rho_{LE}$, and replacing the sum 
$\sum_i {\rm ln}\, p(\rho_i|R)$ by the single term ${\rm ln}\, p(\rho_{LE}|R)$.   For the case of interest here, where $\rho_{LE}$ is (at least) a few times larger than $\rho_1$ and $R_t$ is of order $1$, we can neglect the noise contribution $p_n(\rho)$. Then the integrals in Eq.~(\ref{eq:rb1}) become especially simple, and yield the estimate:
\be\label{eq:rb3}
R_B \approx 
\begin{cases}
(\rho_{LE}/\rho_1)^3 & \text{if $p(R) \propto 1/R$} \\
2\, (\rho_{LE}/\rho_1)^3 & \text{if $p(R) =$ const \, .} 
\end{cases}
\ee

If instead we set the lower integration limit to $\rho_{th} = \rho_1$, then the $p_n(\rho)$
contribution to $p(\rho)$ can no longer be neglected.  But the integrals are still 
straightforward, and we find:
 \be\label{eq:rb3}
R_B \approx 
\begin{cases}
1 \ \ \  & \text{if $p(R) \propto 1/R$} \\
2  \ \ \ & \text{if $p(R) =$ const \, .} 
\end{cases}
\ee
\noindent Thus (as emphasized above) simply incorporating the information that there are no events {\it between} $\rho_1$ and  $\rho_{LE}$ results in a far beter estimate.
